Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex
Progress on the background section - great examples guys\!

%
% LaTeX template for prepartion of submissions to PLDI'16
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'16 web site.
% 
\documentclass[nocopyrightspace,10pt]{sigplanconf}
% \documentclass[pldi-cameraready]{sigplanconf-pldi16}

@@

\section{Background and Examples}\label{sec:background}

@@

\ADComment{Points to make are: there is spinning, either on mutex
implementation or main loop, so fairness is required.  Note OpenCL
features: kernel keyword, global qualifier, ``get'' functions.  Note
that pop\_or\_steal is executed by all, but internally the master
manipulates the queue.  t is a single task, but is chunky enough that
all threads can process part of it.}

@@

An OpenCL program is divided into \emph{host} and \emph{device}
components.  A host application runs on the CPU and launches one or
more \emph{kernels} that run on accelerator devices---GPUs in the
context of this paper.  A kernel is writen in OpenCL C, based on C99.
All threads executing a kernel execute the same entry function with
identical arguments, but a thread may query its unique thread id in
order to access different data or follow different control flow paths
to other threads.

\paragraph{Work stealing example}
  \ADComment{Describe some
  aspects of the example, saying that the mutex is not a primitive,
  and can be coded up using the memory model but is not safe due to
  the execution model.  Make it clear that the lock function uses
  busy-waiting.}  
Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue.
%
Work stealing is useful when a computation can be split into multiple
tasks that can be processed independently, but where the number of
tasks is not known ahead of time because processing a given task may
create an arbitrary number of new tasks. Work stealing enables dynamic
balancing of tasks across several processing units, OpenCL workgroups
in our case. Each workgroup has an associated task pool from which it
obtains tasks to process, and to which it stores newly created
tasks. When a workgroup's pool becomes empty, a workgroup---rather
than terminating---tries to \emph{steal} a task from an other
pool. Hence, workgroups keep on processing tasks until all the work is
done.

\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation of
a work stealing application. All threads start their execution at the
function marked with the $\mathsf{kernel}$ keyword. They receive a
pointer to the task pools that reside in global memory, and which were
initialized by the host program beforehand to contain the initial
tasks. Each thread can identify the workgroup it belongs to via the
function $\getgroupid$, and OpenCL provides a series of primitives in the
same vein to access for instance the thread global id, or its local id
within a workgroup. Here the group id is stored into the variable
$\mathsf(pool\_id)$ (which is, by default, in private memory) and used
to access the relevant task pool. The $\mathsf(pop\_or\_steal)$ function
is called by all threads of a workgroup, although inside only the thread
with a local id equal to $0$, also called \emph{master thread} of the
workgroup, will try to pop or steal a task from the pools. The master
thread uses local memory inside $\mathsf(pop\_or\_steal)$ to ensure that
the returned value is valid for all threads of its workgroup. If a task
was obtained, then all threads of the workgroup can participate in the
processing of the task inside the $\mathsf{process\_task}$
function. \HEComment{Manage to say that a task is considered big enough
to benefit from parallel computation} If some new tasks are created
during the processing, it is again the master thread who will be in
charge of pushing them to the pool. Although not depicted here,
concurrent accesses to each pool is guarded by a mutex which is
implemented by atomic compare and swap operations on global
memory. Henceforth, master threads may be spinning on mutexes, and
workgroups are also spinning on the main loop, so a fair scheduling of
workgroups is desirable.

@@

\paragraph{Graph traversal} Figure~\ref{fig:graphsearch} shows the
skeleton of a frontier based graph traversal algorithm (this approach
has been shown to be efficient on GPUs in~\ref{...}). The kernel is
given three arguments in global memory: a graph structure, and two
arrays that can contain graph nodes. Initially, {\tt n0} contains the
starting nodes to process. Threads keep three local variables: {\tt
  level} which indicates the frontier level which is currently being
processed, {\tt in\_nodes} and {\tt out\_nodes} which point to
distinct node arrays. {\tt in\_nodes} contains the nodes to be
processed for the current frontier, while {\tt out\_nodes} contains
the nodes to be processed during the next frontier.

@@

The threads executing a kernel are hierarchically organised.  A kernel
is executed by a number of \emph{workgroups}, and each workgroup
consists of a fixed number of \emph{threads}.  Within a workgroup,
threads may also be organised into \emph{subgroups} (termed
\emph{warps} and \emph{wavefronts} on NVIDIA and AMD architectures,
respectively).

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups \TSComment{This should explicitly say \emph{workgroups
    executing the same kernel}, otherwise HSA has some weirdness},
stating~\cite{...}: \emph{``A conforming implementation may choose to
  serialize the workgroups so a correct algorithm cannot assume that
  workgroups will execute in parallel.  There is no safe and portable
  way to synchronize across the independent execution of workgroups
  since once in the work-pool, they can execute in any order.''}
OpenCL is not alone in this regard: CUDA similarly provides no
guarantees~\cite{...}.

HSA provides limited, one-way guarantees,
stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
  for values written by work-group B without deadlock provided ... A
  comes after B in work-group flattened ID order}. However, this
one-way communication is not sufficient so support either the
work stealing or graph application examples of
Figure~\ref{fig:graphsearch} and \ref{fig:workstealing}. As described
above, both require \emph{symmetric} communication between
threads. \ADComment{Tyler, can you state the HSA guarantee?},
\TSComment{How is this?}~\cite{...}.

@@

\paragraph{Identity and size functions}

Builtin functions $\getlocalid()$ and $\getglobalid()$ provide a
thread with an id that is unique within its workgroup and unique
across all workgroups, respectively.  Each workgroup has an id,
obtained via $\getgroupid()$.  The number of threads per workgroup and
the number of workgroups are obtained via $\getlocalsize()$ and
$\getnumgroups()$.  A thread's local and global id are related via the
equation $\getglobalid() = \getgroupid() \times \getlocalsize() +
\getlocalid()$.

In general, an OpenCL kernel is organised as a multi-dimensional grid
of multi-dimensional workgroups, and the above functions each take
an integer dimension argument.  For simplicity, we consider only the
one-dimensional case throughout the paper, which captures all the
GPU-accelerated irregular algorithms we are aware of, and use
e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.

\paragraph{Memory model}

An OpenCL kernel has access to four memory spaces.  Shared virtual
memory (SVM) is accessible to all threads, and to the host
application.  Atomic operations on this memory space enable
fine-grained communication between host and device.  Our
implementation of cooperative kernels
(Section~\ref{sec:implementation}) depends on this recent OpenCL
feature.  Global memory is shared among all threads executing a
kernel, but can only be safely accessed by the host application before
or after kernel execution.  Local memory is shared among all threads
in the same workgroup: each workgroup has its own portion of local
memory, which can be used for fast intra-workgroup communication.
Each thread has an allocation of very fast private memory, where
function-local variables are allocated.

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{...}.
\ADComment{Perhaps say more.}  \ADComment{Can we illustrate using the
  examples?}

>>>>

%
% LaTeX template for prepartion of submissions to PLDI'16
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'16 web site.
% 
\documentclass[numbers,nocopyrightspace,10pt]{sigplanconf}
% \documentclass[pldi-cameraready]{sigplanconf-pldi16}

@@

\section{Background and Motivating Examples}\label{sec:background}

@@

@@

An OpenCL program is divided into \emph{host} and \emph{device}
components.  A host application runs on the CPU and launches one or
more \emph{kernels} that run on accelerator devices---GPUs in the
context of this paper.  A kernel is writen in OpenCL C, based on C99.
All threads executing a kernel execute the same entry function with
identical arguments, but a thread may invoke the $\getglobalid()$
function to query its unique thread id in order to access different
data or follow different control flow paths to other threads.

The threads executing a kernel are hierarchically organised.  A kernel
is executed by a number of \emph{workgroups}, and each workgroup
consists of a fixed number of threads.  Builtin functions
$\getlocalid()$ and $\getgroupid()$ return a thread's local id within
its workgroup and the id of the workgroup, respectively.\footnote{In general, an OpenCL kernel is organised as a multi-dimensional grid
of multi-dimensional workgroups, and the above functions each take
an integer dimension argument.  For simplicity, we consider only the
one-dimensional case throughout the paper, which captures all the
GPU-accelerated irregular algorithms we are aware of, and use
e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
}  The number
of threads per workgroup and the number of workgroups are obtained via
$\getlocalsize()$ and $\getnumgroups()$.  A thread's local and global
id are related via the equation $\getglobalid() = \getgroupid() \times
\getlocalsize() + \getlocalid()$.

\ADComment{Maybe cut?} Within a workgroup, threads may also be
organised into \emph{subgroups} (termed \emph{warps} and
\emph{wavefronts} on NVIDIA and AMD architectures, respectively).

\paragraph{Memory spaces and memory model}
An OpenCL kernel has access to four memory spaces.  Shared virtual
memory (SVM) is accessible to all threads, and to the host
application.  Global memory is shared among all threads executing a
kernel, but can only be safely accessed by the host application before
or after kernel execution.  Local memory is shared among all threads
in the same workgroup: each workgroup has its own portion of local
memory, which can be used for fast intra-workgroup communication.
Each thread has an allocation of very fast private memory, where
function-local variables are allocated.

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{...}.  In particular,
atomic SVM operations enable fine-grained communication between host
and device.  Our implementation of cooperative kernels
(Section~\ref{sec:implementation}) depends on this recent OpenCL
feature.

\paragraph{Work stealing example}
Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue.
%
The work stealing paradigm is useful when a computation can be split into multiple
tasks that can be processed independently, but where the number of
tasks is not known ahead of time because processing a given task may
create an arbitrary number of new tasks. Work stealing enables dynamic
balancing of tasks across several processing units, OpenCL workgroups
in our case. Each workgroup has an associated task pool from which it
obtains tasks to process, and to which it stores newly created
tasks. When a workgroup's pool becomes empty, a workgroup---rather
than terminating---tries to \emph{steal} a task from an other
pool. Hence, workgroups keep on processing tasks until all the work is
done.

\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation
of a work stealing algorithm. All threads start their execution at the
function marked with the $\mathsf{kernel}$ keyword. They receive a
pointer to the task pools, which reside in global memory, and which
were initialized by the host program beforehand to contain the initial
tasks. Each thread can identify the workgroup it belongs to via the
function $\getgroupid()$.  Here the group id is stored into the
variable $\mathsf{pool\_id}$ (which is, by default, in private memory)
and used to access the relevant task pool. The
$\mathsf{pop\_or\_steal}$ function is called by all threads of a
workgroup, although inside only the thread with a local id equal to
$0$, also called the \emph{master thread} of the workgroup,
\ADComment{Hugues, I changed; please check:} will try to pop a task
from the workgroup's pool, or steal a task from a different pool. The
master thread uses local memory inside $\mathsf{pop\_or\_steal}$ to
ensure that the returned value is valid for all threads of its
workgroup. If a task was obtained, then all threads of the workgroup
participate in the processing of the task inside the
$\mathsf{process\_task}$ function. \HEComment{Manage to say that a
  task is considered big enough to benefit from parallel computation}
\ADComment{I think that would be too much detail given the space we
  have, so I would not delve into this. Remove both comments if you
  agree.} Processing this task may lead to the creation of further
tasks, which the master thread will push to the workgroup's pool.

Although not depicted here, concurrent accesses to each pool inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex which is implemented using atomic compare and swap operations on
global memory.  As a result, the kernel presents two opportunities for
spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task to process.  The algorithm thus depends
on a fair scheduler, which, as discussed in Section~\ref{sec:intro},
is not provided by current GPU programming models.  In
Section~\ref{sec:cooperativeexamples} we show how the example can be
updated to use our cooperative threads programming model to resolve
the scheduling issue.

@@

\paragraph{Graph traversal example} Figure~\ref{fig:graphsearch} shows the
skeleton of a frontier based graph traversal algorithm (this approach
has been shown to be efficient on GPUs in~\ref{...}). The kernel is
given three arguments in global memory: a graph structure, and two
arrays that can contain graph nodes. Initially, {\tt n0} contains the
starting nodes to process. Threads keep three local variables: {\tt
  level} which indicates the frontier level which is currently being
processed, {\tt in\_nodes} and {\tt out\_nodes} which point to
distinct node arrays. {\tt in\_nodes} contains the nodes to be
processed for the current frontier, while {\tt out\_nodes} contains
the nodes to be processed during the next frontier.

@@

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups executing the same kernel, stating~\cite{...}: \emph{``A
  conforming implementation may choose to serialize the workgroups so
  a correct algorithm cannot assume that workgroups will execute in
  parallel.  There is no safe and portable way to synchronize across
  the independent execution of workgroups since once in the work-pool,
  they can execute in any order.''}  OpenCL is not alone in this
regard: CUDA similarly provides no guarantees~\cite{...}.

HSA provides limited, one-way guarantees,
stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
  for values written by work-group B without deadlock provided ... A
  comes after B in work-group flattened ID order}. However, this
one-way communication is not sufficient so support either the
work stealing or graph application examples of
Figures~\ref{fig:workstealing} and \ref{fig:graphsearch}. As described
above, both require \emph{symmetric} communication between
threads. \ADComment{Tyler, can you state the HSA guarantee?},
\TSComment{How is this?}~\cite{...}.

@@
